### 2014-04-06 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
adding copper mountain multilevel slides

Recent work in Monte Carlo methods for problems in finance and
stochastic partial differential equations has indicated that the
computationally complexity of the problem can be dramatically reduced
by incorporating multigrid concepts into the solution scheme
\cite{heinrich_2001,giles_2008,cliffe_2011}. In this work, we adapt
those ideas and apply them to the Monte Carlo problem for linear
systems as a means of reducing the computational complexity of the
algorithm. To begin, we first introduce the elliptic model problem for
our numerical experiments and compare the spectral behavior of the
Monte Carlo method to traditional iterative smoothers that would be
used with the multigrid method. Next, we present the multilevel Monte
Carlo method for linear systems using a general algebraic
formulation. Finally, we present results using the model problem the
demonstrate the superiority of the multilevel method in terms of time
to solution.

@@

Using this idea we then define a \textit{prolongation operator},
$\ve{P}_{l+1}^l$, which maps a vector defined on grid $l+1$ to a
vector defined on grid $l$ and a \textit{restriction operator},
$\ve{R}_l^{l+1}$, which maps a vector defined on grid $l$ to a
vector defined on grid $l+1$. We can then define the following
equivalent expectation value for a given level $l$:
\begin{equation}
  E(\ve{x}_{l} - \ve{x}_{(l+1)}) = \Big(\ve{I} - \ve{P}_{l+1}^l
  \ve{R}_l^{l+1}\Big) \hat{\ve{x}}_{l}\:,
  \label{eq:level_expectation}
\end{equation}
with $\hat{\ve{x}}_{l}$ given by Eq~(\ref{eq:standard_estimator}) as
the solution on grid $l$. In this estimator, the action of the
restriction operator maps the tally to the coarse level to compute the
observations on that grid and the action of the prolongation operator
on the coarse tally maps the result back to the fine level. Using this
estimator, Algorithm~\ref{alg:mlamc} gives the multilevel Monte Carlo
algorithm for linear systems now with an arbitrary right hand side
vector, $\ve{b}$.
\begin{algorithm}[h!]
  \caption{Multilevel Adjoint Monte Carlo}
  \label{alg:mlamc}
  \begin{algorithmic}[1]

    \For{ l = 0...L }
    \State $\ve{A}_l \ve{x}_l = \ve{b}_l$
    \Comment{Solve the $l^{th}$ level problem with adjoint Monte
      Carlo}
    \If{ l != L }
    \State $\ve{x}_l = (\ve{I} - \ve{P}_{l+1}^l\ve{R}_l^{l+1}) \ve{x}_{l}$
    \Comment{Apply the multilevel tally}
    \EndIf
    \EndFor

    \For{ l = L...1 }
    \State $\ve{x}_{l-1} = ( \ve{I} + \ve{P}_{l}^{l-1} ) \ve{x}_{l}$
    \Comment{Collapse the tallies to the finest grid}
    \EndFor

  \end{algorithmic}
\end{algorithm}
This algorithm permits each of the levels to be computed independently
and then combined appropriately with the prolongation and restriction
operators. Like other multigrid algorithms for linear systems, we are
required to produce a representation of the linear operator on each
level, $\ve{A}^l$, as well as the right hand side, $\ve{b}^l$. The
formulation of Algorithm~\ref{alg:mlamc} places no restrictions on the
form of $\ve{A}^l$, $\ve{b}^l$, $\ve{P}_{l+1}^l$, or $\ve{R}_l^{l+1}$
and therefore could be used with any multigrid scheme including those
arising from algebraic multigrid formulations.

%%---------------------------------------------------------------------------%%
\section{Results}
\label{sec:results}
To demonstrate the multilevel Monte Carlo algorithm we solve the
Poisson problem presented in \S~\ref{sec:fourier_analysis}, this time
with a grid of size $J = 1024$. For the coarsening parameter we will
first use a value of $M = 2$. For simplicity, we will assume a basic
geometric multigrid formulation of the linear problem per
\cite{briggs_multigrid} with the the following prolongation operator
for $M=2$:
\begin{subequations}
  \begin{gather}
    x^h_{2j} = x_j^{2h} \\
    x^h_{2j+1} = \frac{1}{2} ( x_j^{2h} + x_{j+1}^{2h} )\:,
  \end{gather}
  \label{eq:m2_prolongation}
\end{subequations}
for $0 \leq j \leq \frac{J}{2} - 1$ and the restriction operator:
\begin{equation}
  x^{2h}_j = \frac{1}{4} ( x_{2j-1}^{h} + 2 x_{2j}^{2h} + x_{2j+1}^{2h} )
  \label{eq:m2_restriction}
\end{equation}
for $1 \leq j \leq \frac{J}{2} - 1$.

@@

>>>>

Recent work in Monte Carlo methods for problems in finance and
stochastic partial differential equations has indicated that the
computationally complexity of the problem can be dramatically reduced
by incorporating multigrid concepts into the solution scheme
\cite{heinrich_2001,giles_2008,cliffe_2011}. In this work, we adapt
those ideas and apply them to the Monte Carlo problem for linear
systems as a means of reducing the computational complexity of the
algorithm. To begin, we first introduce the elliptic model problem for
our numerical experiments and compare the spectral behavior of the
Monte Carlo method to traditional iterative smoothers that would be
used with the multigrid method. Next, we present the multilevel Monte
Carlo method for linear systems using a general algebraic
formulation. Finally, we present results using the model problem the
demonstrate the superiority of the multilevel method in terms of time
to solution using both geometric and algebraic multigrid
representations of the linear problem.

@@

Using this idea we then define a \textit{prolongation operator},
$\ve{P}_l$, which maps a vector defined on grid $l+1$ to a
vector defined on grid $l$ and a \textit{restriction operator},
$\ve{R}_l$, which maps a vector defined on grid $l$ to a
vector defined on grid $l+1$. We can then define the following
equivalent expectation value for a given level $l$:
\begin{equation}
  E(\ve{x}_{l} - \ve{x}_{l+1}) = \Big(\ve{I} - \ve{P}_l
  \ve{R}_l\Big) \hat{\ve{x}}_{l}\:,
  \label{eq:level_expectation}
\end{equation}
with $\hat{\ve{x}}_{l}$ given by Eq~(\ref{eq:standard_estimator}) as
the solution on grid $l$. In this estimator, the action of the
restriction operator maps the tally to the coarse level to compute the
observations on that grid and the action of the prolongation operator
on the coarse tally maps the result back to the fine level. Using this
estimator, Algorithm~\ref{alg:mlamc} gives the multilevel Monte Carlo
algorithm for linear systems now with an arbitrary right hand side
vector, $\ve{b}$, and an initial solution guess, $\ve{x}^0$.
\begin{algorithm}[h!]
  \caption{Multilevel Monte Carlo Method}
  \label{alg:mlamc}
  \begin{algorithmic}[1]
    \For{ l = 0...L }
    \State $\ve{P}_l = P(\ve{A}_l)$
    \Comment{Build the prolongation and restriction operators for
      the $l^{th}$ level.}
    \State $\ve{R}_l = c \ve{P}_l^T$
    \State $\ve{r}_l = \ve{b}_l - \ve{A}_l \ve{x}_l^0$
    \Comment{Build the $l^{th}$ level residual.}
    \State $\ve{d}_l = \hat{\ve{A}}_l^{-1} \ve{r}_l$
    \Comment{Solve the $l^{th}$ level problem with adjoint Monte
      Carlo}
    \If{ l != L }
    \State $\ve{d}_l = (\ve{I} - \ve{P}_l\ve{R}_l) \ve{d}_{l}$
    \Comment{Apply the multilevel tally}
    \State $\ve{A}_{l+1} = \ve{R}_l \ve{A}_l \ve{P}_l$
    \Comment{Construct the next level.}
    \State $\ve{x}_{l+1}^0 = \ve{R}_l \ve{x}_l^0$
    \State $\ve{b}_{l+1} = \ve{R}_l \ve{b}_l$
    \EndIf
    \EndFor

    \For{ l = L...1 }
    \State $\ve{d}_{l-1} = ( \ve{I} + \ve{P}_{l} ) \ve{d}_{l}$
    \Comment{Collapse the tallies to the finest grid}
    \EndFor

    \State $\ve{x} = \ve{x}^0 + \ve{d}_0$

  \end{algorithmic}
\end{algorithm}
This algorithm permits each of the levels to be computed independently
and then combined appropriately with the prolongation and restriction
operators. Like other multigrid algorithms for linear systems, we are
required to produce a representation of the linear operator on each
level, $\ve{A}_l$, as well as the right hand side, $\ve{b}_l$. The
formulation of Algorithm~\ref{alg:mlamc} places no restrictions on the
form of $\ve{A}_l$, $\ve{b}_l$, $\ve{P}_l$, or $\ve{R}_l$
and therefore could be used with any multigrid scheme including those
arising from algebraic multigrid formulations.

%%---------------------------------------------------------------------------%%
\section{Geometric Multigrid Results}
\label{sec:geometric_results}
To demonstrate the multilevel Monte Carlo algorithm we solve the
Poisson problem presented in \S~\ref{sec:fourier_analysis}, this time
with a grid of size $J = 1024$. For the coarsening parameter we will
first use a value of $M = 2$. For simplicity, we will assume a basic
geometric multigrid formulation of the linear problem per
\cite{briggs_multigrid} with the the following prolongation operator
for $M=2$:
\begin{subequations}
  \begin{gather}
    x^h_{2j} = x_j^{2h} \\
    x^h_{2j+1} = \frac{1}{2} ( x_j^{2h} + x_{j+1}^{2h} )\:,
  \end{gather}
  \label{eq:m2_prolongation}
\end{subequations}
for $0 \leq j \leq \frac{J}{2} - 1$ and the restriction operator:
\begin{equation}
  x^{2h}_j = \frac{1}{4} ( x_{2j-1}^{h} + 2 x_{2j}^{2h} + x_{2j+1}^{2h} )
  \label{eq:m2_restriction}
\end{equation}
for $1 \leq j \leq \frac{J}{2} - 1$.

@@

%%---------------------------------------------------------------------------%%
\section{Algebraic Multigrid Results}
\label{sec:algebraic_results}
